Renewal and stability in populations structured by remaining years of life

### 2016-10-28 ThanoRepro/ThanoRepro.tex
minor edits to get the ball rolling in anticipation of having some empirical fertility rates by years lived and left

The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing
the long term consequences of a set of vital rates. All lines
of contemporary demographic research are relatable to the renewal model, and by
extention to stable population theory. The purpose of this paper is to describe
the Lotka-Leslie renewal model when structured by years of life left.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
%To present the renewal model based on thanatological age is not necessarily a
%suggestion for a better way to model or project population. Rather, 
The years left renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here provide a broader treatment allowing for
growth in the population.
The years lived renewal model and the years left renewal
model both describe the movement of population stocks over time, but from opposite vantage points. %Specifying a model of thanatological renewal expands the toolbox available to approach complex temporal population structure.

@@

\section*{Requisites of a years left renewal model}
The Lotka-Leslie renewal model \citep{leslie1945use,sharpe1911problem} is
situated in a closed, unisex, homogenous population framework. With
parameters equated, the continuous form of the model contains a component for
survival, $\ell(a)$, a scaling factor for growth, $e^{-ra}$, and a component for fertility, $m(a)$:

@@

Each of these elements is structured by
age, $a$. The model can be approximated by observed data because
age is typically known.% The survival function, $\ell(a)$, derives
%from a lifetable, usually based on age-specific event-exposure mortality rates,
%which require counts of decedents and an estimate of the
%population exposed to risk in the same chronological age classes.
Age in the growth factor, $e^{-ra}$, pertains to the passage of
time more than it does to chronological age (these are synchronous for birth
cohorts). The product of this growth factor and $\ell(a)$ gives a
function proportional to the stable age structure, $c(a)$, and it is often
intuitive to imagine these two elements together as one for this reason. Female
fertility rates, $m(a)$, are based on daughters born to mothers in this limited
single-sex model. This model lends itself to commonly collected data because age
is typically known, and can be asked.
% For the two essential rate componenents underlying the renewal
 % equation~\eqref{eq:lotka}, classification by chronological age is essentially given--- it is deterministic in the sense that time passed since birth is typically known by individuals. The kinds of errors in such classification are mostly due to innumeracy, rounding, exaggeration, and similar kinds of biases, but even in such cases, chronological age is fundamentally knowable.

>>>>

The renewal model relates
fertility, mortality, and population structure in a parsimonious form, revealing
the long term consequences of a set of vital rates. All lines
of contemporary demographic research are relatable to the renewal model, and by
extention to stable population theory. The purpose of this paper is to describe
the Lotka-Leslie renewal model when structured by years of life left.
%Some might say that stable population
%theory is not relevant to everyday demography, since populations are always
% changing. Even if change precludes the notion of stability as a prediction, models of long-run consequnces still give a
%useful measure of the demographic force implied by a given set of rates.
%In this and other ways, stability is never far away. That population
% projections lead into a stable state means that some form of stability is implied by the routine practice of demography.
%Stability or even stationarity are imaginable for the world population of
% humans one day. Models of stationary populations are easily applied by analogy to
%closed or nearly-closed birth cohorts. Finally, stable population theory is
%relavant for modelling some non-human populations. 
%We have stated that many lifecourse transitions covary more with remaining
%lifetime than with life lived. In an even more general sense, by thinking about
% remaining lifetime as a particular class of time-to-transition, one may generalize the lifetable methods presented here into new area applications, such as education, partnering, the economic lifecycle, or migration.
%To present the renewal model based on thanatological age is not necessarily a
%suggestion for a better way to model or project population. Rather, 
The years-left renewal model represents an alternative and coherent way to
describe the process of population renewal by relating renewal to a death flow
rather than to a birth flow. I have described a limited case of this
perspective on renewal for the case of stationary populations
\citep{riffe2015force}, and here provide a broader treatment allowing for
growth in the population.
The years-lived renewal model and the years-left renewal
model both describe the movement of population stocks over time, but from opposite vantage points. %Specifying a model of thanatological renewal expands the toolbox available to approach complex temporal population structure.

@@

\section*{Requisites of a years-left renewal model}
The Lotka-Leslie renewal model \citep{sharpe1911problem, leslie1945use} is
situated in a closed, unisex, homogenous population framework. With
parameters equated, the continuous form of the model contains a component for
survival, $\ell(a)$, a scaling factor for growth, $e^{-ra}$, and a component for fertility, $m(a)$:

@@

Each of these elements is structured by
age, $a$. The model can be approximated by observed data because
age is typically known. % The survival function, $\ell(a)$, derives
%from a lifetable, usually based on age-specific event-exposure mortality rates,
%which require counts of decedents and an estimate of the
%population exposed to risk in the same chronological age classes.
Age in the growth factor, $e^{-ra}$, pertains to the passage of
time more than it does to chronological age (this is synchronous with birth
cohorts). The product of this growth factor and $\ell(a)$ gives a
function proportional to the stable age structure, $c(a)$, and it is often
intuitive to imagine these two elements together as one for this reason. Female
fertility rates, $m(a)$, are based on daughters born to mothers in this limited
single-sex model. The functions $m(a)$ and $\ell(a)$
together intuitively make for net female fertility, and their age-wise product
is sometimes written as $\phi(a)$. This model lends itself to commonly collected data because age
is typically known, and can be asked.
% For the two essential rate componenents underlying the renewal
 % equation~\eqref{eq:lotka}, classification by chronological age is essentially given--- it is deterministic in the sense that time passed since birth is typically known by individuals. The kinds of errors in such classification are mostly due to innumeracy, rounding, exaggeration, and similar kinds of biases, but even in such cases, chronological age is fundamentally knowable.
